---
title: "Assignment 10: Data Scraping"
author: "Liam Connolly"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A10_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Check your working directory

```{r 1. Project Setup, message = FALSE}
#1 
library(tidyverse)
library(rvest)
setwd("/home/guest/EDE_Fall2025/Data/Assignment10")
getwd()
```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2024 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2024>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2 
url <- "https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2024"
page <- read_html(url)
```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PWSID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Day Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings)".

```{r scrape.the.data}
#3 
system_info <- page %>% 
  html_elements("table") %>% 
  .[[1]] %>% 
  html_table(fill = TRUE)
water_system_name <- system_info[1,2]
pwsid             <- system_info[2,2]
ownership         <- system_info[3,2]
tables <- page %>% html_elements("table")
max_day_table <- tables[[13]] %>% html_table(fill = TRUE)
print(names(max_day_table))
print(head(max_day_table))
max_day_use <- max_day_table$`Max DayUse (MGD)`
str(max_day_use)
max_day_use
```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc... 

5. Create a line plot of the maximum daily withdrawals across the months for 2024, making sure, the months are presented in proper sequence.

```{r create.a.dataframe.from.scraped.data}
#4 
tables <- page %>% html_elements("table") %>% map(html_table, fill = TRUE)
four_row_tables <- keep(tables, ~ nrow(.x) == 4)
max_day_table <- four_row_tables[[which(sapply(four_row_tables, ncol) == 9)]]
max_long <- tibble(
  Month = c(max_day_table[[1]], max_day_table[[4]], max_day_table[[7]]),
  MaxDay_MGD = c(max_day_table[[3]], max_day_table[[6]], max_day_table[[9]])
)
df <- tibble(
  Water_System = rep(water_system_name, 12),
  PWSID        = rep(pwsid, 12),
  Ownership    = rep(ownership, 12),
  Month        = max_long$Month,
  MaxDay_MGD   = max_long$MaxDay_MGD,
  Year         = 2024
) %>%
  mutate(
    # Converts abbreviated month names to proper dates
    Date = as.Date(paste(Month, Year, "1"), format = "%b %Y %d")
  )
#5 
df %>%
  mutate(Month = factor(Month, levels = month.abb)) %>%
  ggplot(aes(x = Month, y = MaxDay_MGD, group = 1)) +
  geom_line(color = "blue", linewidth = 1.2) +
  geom_point(size = 3) +
  labs(
    title = "Durham 2024 Maximum Daily Withdrawals",
    x = "Month",
    y = "Max Daily Use (MGD)"
  ) +
  theme_minimal()
```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function with two inputs - "PWSID" and "year" - that:
  - Creates a URL pointing to the LWSP for that PWSID for the given year
  - Creates a website object and scrapes the data from that object (just as you did above)
  - Constructs a dataframe from the scraped data, mostly as you did above, but includes the PWSID and year provided as function inputs in the dataframe. 
  - Returns the dataframe as the function's output


```{r construct.a.scraping.function}
#6. 
scrape_LWSP <- function(PWSID, year) {
  url <- paste0("https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=", 
                PWSID, "&year=", year)
  page <- read_html(url)
  system_name <- "Durham"         
  ownership   <- "Municipality"
  tables <- page %>% html_elements("table") %>% map(html_table, fill = TRUE)
  max_day_table <- keep(tables, ~ any(grepl("Max Day", names(.x), 
                                            ignore.case = TRUE)))[[1]]
  text_block <- max_day_table %>% as.character() %>% paste(collapse = " ")
  maxday <- stringr::str_extract_all(text_block, 
                        "\\d+\\.?\\d*")[[1]] %>% as.numeric()
  maxday <- maxday[1:12]
  months <- month.abb
  tibble(
    Water_System = rep(system_name, 12),
    PWSID = rep(PWSID, 12),
    Ownership = rep(ownership, 12),
    Month = factor(months, levels = month.abb),
    MaxDay_MGD = maxday,
    Year = rep(year, 12),
    Date = as.Date(paste(year, months, "1"), format = "%Y %b %d")
  )
}
#I am completely befuddled by this, maybe this works???
```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2020. Then show the structure of the dataframe with either the `str()` or the `glimpse()` function. 

```{r fetch.and.plot.Durham.2020.data}
#7 
durham2020 <- scrape_LWSP("03-32-010", 2020)
ggplot(durham2020, aes(x = Month, y = MaxDay_MGD, group = 1)) +
  geom_line(linewidth = 1.2, color = "blue") +
  geom_point(size = 2) +
  theme_minimal() +
  labs(title = "Durham Max Daily Water Withdrawal (2020)",
       x = "Month", y = "Max Day Use (MGD)")
glimpse(durham2020)
```

8. Use the function above to extract data for Cary (PWSID = '03-92-020') in 2020. Combine this data with the Durham data collected above and create a plot that compares Cary's to Durham's water withdrawals.

```{r fetch.and.plot.Cary.2020.data}
#8 
cary2020 <- scrape_LWSP("03-92-020", 2020)
both2020 <- bind_rows(
  durham2020 %>% mutate(System = "Durham"),
  cary2020 %>% mutate(System = "Cary")
)
ggplot(both2020, aes(x = Month, y = MaxDay_MGD, color = System, 
                     group = System)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  theme_minimal() +
  labs(title = "Cary vs Durham Max Daily Water Withdrawal (2020)",
       x = "Month", y = "Max Day Use (MGD)")
glimpse(both2020)
```


9. Use the code & function you created above to plot Cary's max daily withdrawal by months for the years 2018 thru 2023.Add a smoothed line to the plot (method = 'loess').

> TIP: See Section 3.2 in the "10_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs. Pipe the output of the map2() function to `bind_rows()` to combine the dataframes into a single one, and use that to construct your plot.

```{r fetch.and.plot.Cary.multiyear.data }
#9 
years <- 2018:2023
cary_multi <- map(years, ~ scrape_LWSP("03-92-020", .x)) %>% bind_rows()
ggplot(cary_multi, aes(x = Date, y = MaxDay_MGD)) +
  geom_line(group = 1, linewidth = 1, color = "green") +
  geom_point(size = 2) +
  geom_smooth(method = "loess", color = "red", se = FALSE) +
  theme_minimal() +
  labs(title = "Cary Max Daily Withdrawal (2018–2023)",
       x = "Date", y = "Max Day Use (MGD)")
```

Question: Just by looking at the plot (i.e. not running statistics), does Cary have a trend in water usage over time?
> Answer: Cary's water usage shows a slight increase over the years. 
There are monthly fluctuations but, the pattern itself is that peak
water demand is growing gradually. 
>
