---
title: "Assignment 8: Time Series Analysis"
author: "Liam Connolly"
date: "Fall 2025"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on
generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A08_TimeSeries.Rmd` (replacing `<FirstLast>`
with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each
instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into
a single PDF file.

## Set up 
1. Set up your session:

* Check your working directory
* Load the tidyverse, lubridate, zoo, and trend packages
* Set your ggplot theme

```{r}
#1
getwd()
setwd("/home/guest/EDE_Fall2025/Data/Raw/Ozone_TimeSeries")
library(tidyverse)
install.packages("tidyverse")
library(lubridate)
install.packages("lubridate")
library(zoo)
install.packages("zoo")
install.packages("trend")
library(trend)
list.files("/home/guest/EDE_Fall2025/Data/Raw/Ozone_TimeSeries")
install.packages("dplyr")
install.packages("purrr")
install.packages("readr")
library(dplyr)
library(purrr)
library(readr)
install.packages("ggplot2")
library(ggplot2)
install.packages("Kendall")
library(Kendall)
```


2. Import the ten datasets from the Ozone_TimeSeries folder in the
Raw data folder. These contain ozone concentrations at Garinger
High School in North Carolina from 2010-2019 
(the EPA air database only allows downloads for one year at a time). 
Import these either individually or in bulk and then combine them
into a single dataframe named `GaringerOzone` of
3589 observation and 20 variables. 

```{r, message = FALSE}
#2
ozone2010 <- read_csv("EPAair_O3_GaringerNC2010_raw.csv") %>%
  mutate(Date = as.Date(Date, format = "%d/%m/%Y"))
ozone2011 <- read_csv("EPAair_O3_GaringerNC2011_raw.csv") %>%
  mutate(Date = as.Date(Date, format = "%d/%m/%Y"))
ozone2012 <- read_csv("EPAair_O3_GaringerNC2012_raw.csv") %>%
  mutate(Date = as.Date(Date, format = "%d/%m/%Y"))
ozone2013 <- read_csv("EPAair_O3_GaringerNC2013_raw.csv") %>%
  mutate(Date = as.Date(Date, format = "%d/%m/%Y"))
ozone2014 <- read_csv("EPAair_O3_GaringerNC2014_raw.csv") %>%
  mutate(Date = as.Date(Date, format = "%d/%m/%Y"))
ozone2015 <- read_csv("EPAair_O3_GaringerNC2015_raw.csv") %>%
  mutate(Date = as.Date(Date, format = "%d/%m/%Y"))
ozone2016 <- read_csv("EPAair_O3_GaringerNC2016_raw.csv") %>%
  mutate(Date = as.Date(Date, format = "%d/%m/%Y"))
ozone2017 <- read_csv("EPAair_O3_GaringerNC2017_raw.csv") %>%
  mutate(Date = as.Date(Date, format = "%d/%m/%Y"))
ozone2018 <- read_csv("EPAair_O3_GaringerNC2018_raw.csv") %>%
  mutate(Date = as.Date(Date, format = "%d/%m/%Y"))
ozone2019 <- read_csv("EPAair_O3_GaringerNC2019_raw.csv") %>%
  mutate(Date = as.Date(Date, format = "%d/%m/%Y"))
ozone2010 <- ozone2010 %>%
  rename(
    Date = `Date`,
    Daily.Max.8.hour.Ozone.Concentration = 
      `Daily Max 8-hour Ozone Concentration`,
    DAILY_AQI_VALUE = `DAILY_AQI_VALUE`
  )
ozone2011 <- ozone2011 %>%
  rename(
    Date = `Date`,
    Daily.Max.8.hour.Ozone.Concentration = 
      `Daily Max 8-hour Ozone Concentration`,
    DAILY_AQI_VALUE = `DAILY_AQI_VALUE`
  )
ozone2012 <- ozone2012 %>%
  rename(
    Date = `Date`,
    Daily.Max.8.hour.Ozone.Concentration = 
      `Daily Max 8-hour Ozone Concentration`,
    DAILY_AQI_VALUE = `DAILY_AQI_VALUE`
  )
ozone2013 <- ozone2013 %>%
  rename(
    Date = `Date`,
    Daily.Max.8.hour.Ozone.Concentration = 
      `Daily Max 8-hour Ozone Concentration`,
    DAILY_AQI_VALUE = `DAILY_AQI_VALUE`
  )
ozone2014 <- ozone2014 %>%
  rename(
    Date = `Date`,
    Daily.Max.8.hour.Ozone.Concentration = 
      `Daily Max 8-hour Ozone Concentration`,
    DAILY_AQI_VALUE = `DAILY_AQI_VALUE`
  )
ozone2015 <- ozone2015 %>%
  rename(
    Date = `Date`,
    Daily.Max.8.hour.Ozone.Concentration = 
      `Daily Max 8-hour Ozone Concentration`,
    DAILY_AQI_VALUE = `DAILY_AQI_VALUE`
  )
ozone2016 <- ozone2016 %>%
  rename(
    Date = `Date`,
    Daily.Max.8.hour.Ozone.Concentration = 
      `Daily Max 8-hour Ozone Concentration`,
    DAILY_AQI_VALUE = `DAILY_AQI_VALUE`
  )
ozone2017 <- ozone2017 %>%
  rename(
    Date = `Date`,
    Daily.Max.8.hour.Ozone.Concentration = 
      `Daily Max 8-hour Ozone Concentration`,
    DAILY_AQI_VALUE = `DAILY_AQI_VALUE`
  )
ozone2018 <- ozone2018 %>%
  rename(
    Date = `Date`,
    Daily.Max.8.hour.Ozone.Concentration = 
      `Daily Max 8-hour Ozone Concentration`,
    DAILY_AQI_VALUE = `DAILY_AQI_VALUE`
  )
ozone2019 <- ozone2019 %>%
  rename(
    Date = `Date`,
    Daily.Max.8.hour.Ozone.Concentration = 
      `Daily Max 8-hour Ozone Concentration`,
    DAILY_AQI_VALUE = `DAILY_AQI_VALUE`
  )
GaringerOzone <- bind_rows(
  ozone2010, ozone2011, ozone2012, ozone2013, ozone2014,
  ozone2015, ozone2016, ozone2017, ozone2018, ozone2019
)

dim(GaringerOzone)
colnames(GaringerOzone)
```

## Wrangle

3. Set your date column as a date class.

4. Wrangle your dataset so that it only contains the columns Date,
Daily.Max.8.hour.Ozone.Concentration, and DAILY_AQI_VALUE. 

5. Notice there are a few days in each year that are missing ozone
concentrations. We want to generate a daily dataset, so we will 
need to fill in any missing days with NA. Create a new data frame
that contains a sequence of dates from 2010-01-01 to
2019-12-31 (hint: `as.data.frame(seq())`).
Call this new data frame Days. Rename the column name in Days to "Date".

6. Use a `left_join` to combine the data frames. Specify the correct order
of data frames within this function so that the final dimensions are
3652 rows and 3 columns. Call your combined data frame GaringerOzone. 

```{r}
# 3
class(GaringerOzone$Date)
summary(GaringerOzone$Date)
GaringerOzone <- GaringerOzone %>%
  mutate(Date = as.Date(Date, format = "%Y-%m-%d"))
# 4
GaringerOzone <- GaringerOzone %>%
  select(Date, Daily.Max.8.hour.Ozone.Concentration, DAILY_AQI_VALUE)
# 5
Days <- data.frame(Date = seq(
  from = as.Date("2010-01-01"),
  to   = as.Date("2019-12-31"),
  by   = "day"
))
# 6
GaringerOzone <- Days %>%
  left_join(GaringerOzone, by = "Date")
```

## Visualize

7. Create a line plot depicting ozone concentrations over time. In this case,
we will plot actual concentrations in ppm, not AQI values.
Format your axes accordingly. Add a smoothed line showing
any linear trend of your data. Does your plot suggest
a trend in ozone concentration over time?

```{r}
#7
GaringerOzone_plot <- GaringerOzone %>%
  mutate(ozone_ppm = Daily.Max.8.hour.Ozone.Concentration / 1000) %>%
  filter(!is.na(Date) & !is.na(ozone_ppm))
ggplot(GaringerOzone_plot, aes(x = Date, y = ozone_ppm)) +
  geom_point(alpha = 0.4, color = "green", size = 1) +  
  geom_line(alpha = 0.3, color = "steelblue") +             
  geom_smooth(method = "lm", se = TRUE, color = "red", linewidth = 1) + 
  labs(
    title = "Daily Maximum 8-hour Ozone Concentration (ppm)",
    subtitle = "Garinger High School, NC — 2010 to 2019",
    x = "Date",
    y = "Ozone concentration (ppm)"
  ) +
  theme_minimal()
```

>Answer: The line plot for daily maximum 8-hour ozone concentrations at 
Garinger High School between 2010-2019 shows a downward-sloping linear trend.
This indicates that ozone levels have generally decreased over this period. 
While there is noticeable seasonal variability, with higher ozone in 
summer and lower in winter, the overall trend suggests air quality has been 
increasing over time.

## Time Series Analysis

Study question: Have ozone concentrations changed over the 2010s 
at this station? 
>Answer: Ozone concentrations at Garinger High School have decreased 
past 2010. The plot of daily maximum 8-hour ozone shows a 
downward-sloping linear trend, indicating a gradual decline. There are some
seasonal variations but, overall the answer is yes. 

8. Use a linear interpolation to fill in missing daily data for 
ozone concentration. 
Why didn't we use a piecewise constant or spline interpolation?

```{r}
#8
GaringerOzone <- GaringerOzone %>%
  arrange(Date) %>%
  mutate(
    ozone_ppm = Daily.Max.8.hour.Ozone.Concentration / 1000,
    ozone_ppm_filled = na.approx(ozone_ppm, x = Date, na.rm = FALSE, rule = 2)
  )
```

> Answer: A spline interpolation fits a smooth curve through all available 
points. While it may look nice, it can overshoot and produce 
unrealistic values. This may exaggerate trends or create false 
data ie. negative values. Piecewise constant fills any missing values with 
either the previous or next recorded value. This creates abnormalities in 
the data and given that ozone concentrations change daily, 
this would incorrectly display natural variability.


9. Create a new data frame called `GaringerOzone.monthly` that contains
aggregated data: mean ozone concentrations for each month. In your pipe,
you will need to first add columns for year and month to form the groupings.
In a separate line of code, create a new Date column with each month-year
combination being set as the first
day of the month (this is for graphing purposes only)

```{r}
#9
GaringerOzone.monthly <- GaringerOzone %>%
  mutate(
    year = year(Date),
    month = month(Date)
  ) %>%
  group_by(year, month) %>%
  summarise(
    ozone_ppm_mean = mean(ozone_ppm_filled, na.rm = TRUE),
    .groups = "drop"
  )
GaringerOzone.monthly <- GaringerOzone.monthly %>%
  mutate(
    Date = as.Date(paste(year, month, "01", sep = "-"))
  ) %>%
  select(Date, ozone_ppm_mean)
```


10. Generate two time series objects. Name the first `GaringerOzone.daily.ts` 
and base it on the dataframe of daily observations. Name the 
second `GaringerOzone.monthly.ts` and base it on the monthly average
ozone values. Be sure that each specifies the correct start and end
dates and the frequency of the time series.

```{r}
#10
GaringerOzone.daily.ts <- ts(
  data = GaringerOzone$ozone_ppm_filled,   
  start = c(2010, 1),                      
  end = c(2019, 365),                     
  frequency = 365                           
)
GaringerOzone.monthly.ts <- ts(
  data = GaringerOzone.monthly$ozone_ppm_mean,
  start = c(2010, 1),  
  end = c(2019, 12),   
  frequency = 12        
)
```


11. Decompose the daily and the monthly time series objects and plot the
components using the `plot()` function.

```{r}
#11
GaringerOzone.daily.decomp <- decompose(GaringerOzone.daily.ts, 
    type = "additive")
plot(GaringerOzone.daily.decomp, xlab = "Time")
GaringerOzone.monthly.decomp <- decompose(GaringerOzone.monthly.ts, 
    type = "additive")
plot(GaringerOzone.monthly.decomp, xlab = "Time")
```


12. Run a monotonic trend analysis for the monthly Ozone series. In this case
the seasonal Mann-Kendall is most appropriate; why is this?

```{r}
#12
ozone_monthly <- GaringerOzone.monthly$ozone_ppm_mean
mk_result <- SeasonalMannKendall(as.ts(GaringerOzone.monthly$ozone_ppm_mean))
print(mk_result)
```

> Answer: tau = -0.058, 2-sided pvalue =0.3477. The seasonal Mann-Kendall test 
is appropriate for this data as it shows strong seasonality. This being true 
as ozone concentrations are higher in summer and lower in winter. 
A standard version of this test may mistake seasonal fluctuations 
as a trend. The seasonal version will in theory compare the same month
across different years. Given the results, there was no strong evidence
of a monotonic trend in monthly ozone concentrations. The negative tau 
suggests a small decline but the trend itself isn't meaningful.   

13. Create a plot depicting mean monthly ozone concentrations over time,
with both a geom_point and a geom_line layer. Edit your
axis labels accordingly.

```{r}
# 13
ggplot(GaringerOzone.monthly, aes(x = Date, y = ozone_ppm_mean)) +
  geom_point(color = "steelblue", size = 2, alpha = 0.7) +  
  geom_line(color = "steelblue", alpha = 0.5) +             
  labs(
    title = "Mean Monthly Ozone Concentrations (ppm)",
    subtitle = "Garinger High School, NC — 2010 to 2019",
    x = "Date",
    y = "Mean ozone concentration (ppm)"
  ) +
  theme_minimal()
```

14. To accompany your graph, summarize your results in context of the 
research question. Include output from the statistical test in parentheses 
at the end of your sentence. Feel free to use multiple sentences in your
interpretation.

> Answer: There were seasonal patterns evident within the plot of mean 
monthly ozone concentrations at Garinger High School between 2010-2019. 
Higher ozone levels were present in the summer months and lower levels 
occurred in the winter. There also was a small downward trend in ozone 
concentration across the ten years which is not as noticeable as the
seasonal fluctuations. Given the findings from the seasonal 
Mann-Kendall test, there was a small negative tau of -0.058
which is not statistically significant with a 2-sided p of 0.348.
In sum, there is no strong evidence of a consistent long-term
decline in monthly ozone concentrations apart from some seaonal variability. 

15. Subtract the seasonal component from the `GaringerOzone.monthly.ts`.
Hint: Look at how we extracted the series components for the
EnoDischarge on the lesson Rmd file.

16. Run the Mann Kendall test on the non-seasonal Ozone monthly series.
Compare the results with the ones obtained with the Seasonal
Mann Kendall on the complete series.

```{r}
#15
GaringerOzone.monthly.decomp <- decompose(GaringerOzone.monthly.ts,
    type = "additive")
ozone_monthly_deseason <- GaringerOzone.monthly.ts - 
  GaringerOzone.monthly.decomp$seasonal
#16
mk_deseason <- MannKendall(ozone_monthly_deseason)
print(mk_deseason)
```

> Answer: Following the removal of the seasonal component, the non
seasonal Mann-Kendall test had tau = -0.067 with a 2-sided
p value of 0.278. By contrast, the seasonal Mann-Kendall
results were tau = -0.058 and p value of 0.348. Both tests indicated
a small downward trend in ozone concentrations between 2010-2019. The test
without seasons had a stronger negative tau and smaller p-value meaning that
seasonal fluctuations had impacted the trend. It is important to note that
the trend was not statistically significant with either test so there was
not strong evidence of an overall consistent long term decline in
monthly ozone. 
